Missing data: Current practice in football research and recommendations for

### 2020-10-04 paper/paper.Rmd
update main document and references.

Authors: David N Borg(1), Robert Nguyen(2), Nicholas J Tierney(3,4)

@@

**Data availability**

@@

Our survey of 136 articles published on football related topics found
that only about one in 20 papers reported missing data. It is reasonable
to assume that this result is representative of current practice in the wider sports science literature. 
The low proportion of papers reporting missingness could be explained by a lack of awareness and education on missing data (https://bjsm.bmj.com/content/early/2020/08/19/bjsports-2020-102607.abstract), 
similar to other areas of sciences (refs-education and
ecology examples)(2)**references needed**. Below we discuss several practices to avoid in
relation to missingness, and provide recommendations for exploring,
visualising and reporting missing data.

@@

Practice to avoid: *Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. Authors sometimes choose to replace missing 
observations with mean values (e.g., participant or group average) [@refs](https://doi.org/10.1016/j.jsams.2018.05.019, https://doi.org/10.1016/j.jsams.2008.01.009). 
While imputing the mean preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002]. Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered. This is discussed in the recommendations section.

Practice to avoid: *Not evaluating the effect, or choice, of imputation.* The 
method used to impute missing data has the potential to
affect the outcome of a study. It is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation--in terms of the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (**ref circulation article-Nick, possible reference from dave - https://doi.org/10.1191%2F1740774504cn032oa**). This is
discussed below.

@@

(Figure \@ref(fig:missing-overview)) give an overall sense of the extent of missing and complete data, in an simulated injury dataset, from Australian Rules football matches, played in August 2020. The dataset contains observed player id, team, and match venue data, in addition to simulated data on injury incidence, and number of sprints per match, and total sprint distance per match. Researchers might be interested in investigating the relationship between number of sprints and total sprint distance with incidence of injury. **should we include this last sentence?**

@@

```{r missing-overview, fig.cap = "Overviews of missing values in simulated injury data in Australian Football matches for August, 2020. Panel A shows... Panel B shows ... Panel C shows ...", fig.height = 10}

>>>>

Authors:

David N Borg(1)

ORCiD: 0000-0002-0152-571X

Twitter: elborgo9

Robert Nguyen(2)

ORCiD:

Twitter:

Nicholas J Tierney(3,4)

ORCiD:

Twitter:

@@

**Data availability**
The code to reproduce the results and plots in the paper can be accessed at: **Nick to host?**

@@

Our survey of 136 articles published on football related topics found
that only about one in 20 papers reported missing data. It is reasonable
to assume that this result is representative of current practice in the wider sports science literature. 
The low proportion of papers reporting missingness could be explained by 
a lack of awareness and education on missing data [@Sainani2020call], 
similar to other areas of sciences (refs-education and
ecology examples)(2)**references needed**. Below we discuss several practices to avoid in
relation to missingness, and provide recommendations for exploring,
visualising and reporting missing data.

@@

Practice to avoid: *Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. Authors sometimes choose to replace missing 
observations with mean values (e.g., participant or group average) [@Sampson2018; @Cresswell2009]. 
While imputing the mean preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002]. Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered.

Practice to avoid: *Not evaluating the effect, or choice, of imputation.* The 
method used to impute missing data has the potential to
affect the outcome of a study. It is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation--in terms of the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (**ref circulation article-Nick, possible reference from dave - https://doi.org/10.1191%2F1740774504cn032oa**).

@@

(Figure \@ref(fig:missing-overview)) give an overall sense of the extent of missing and complete data, in an simulated injury dataset.

@@

```{r missing-overview, fig.cap = "Overviews of missing values in simulated injury data in Australian Football matches for August, 2020. The dataset contains observed player id, team, and match venue data, in addition to simulated data on injury incidence, and number of sprints per match, and total sprint distance per match. Researchers might be interested in investigating the relationship between number of sprints and total sprint distance with incidence of injury. Panel A shows... Panel B shows ... Panel C shows ... Panel D shows ...", fig.height = 10}
